\documentclass[11pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{tikz}
\usepackage{caption}
\usepackage{subcaption}
\usepackage[hidelinks,
            pdftitle={GPU-Accelerated 3D Point Cloud Histogram Equalization},
            pdfauthor={Shahid Khan}]{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{array}
\usepackage{float}

\lstset{
  language=C++,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  frame=single,
  breaklines=true
}

\title{\textbf{GPU-Accelerated 3D Point Cloud Histogram Equalization}\\
\large COL380 Assignment 2 -- CUDA Implementation and Analysis}
\author{Shahid Khan \\ ee1221163}
\date{}

\begin{document}

% ---- Title page ----
\maketitle
\thispagestyle{empty}
\vfill
\begin{center}
\large Indian Institute of Technology Delhi\\
Department of Computer Science and Engineering\\[0.5em]
\today
\end{center}
\newpage

% ---- Table of contents ----
\thispagestyle{empty}
\tableofcontents
\newpage

% =============================================================================
\section{Introduction}
% =============================================================================

Histogram equalization is a classical image-processing technique that improves
contrast by redistributing intensity values so that the empirical CDF of the
output is approximately uniform.  This assignment extends that idea to a 3-D
point cloud: each point $p_i$ carries an intensity $I_i\in[0,255]$, and its
equalized intensity is computed from the local neighbourhood histogram using
\begin{equation}
      I'_i = \left\lfloor
      \frac{\bigl(\mathrm{CDF}(I_i)-\mathrm{CDF}_{\min}\bigr)\cdot(L-1)}{m-\mathrm{CDF}_{\min}}
      \right\rfloor,\qquad L=256,
\end{equation}
where the CDF counts points (including $p_i$ itself) in the local window with
intensity $\le I_i$, $\mathrm{CDF}_{\min}$ is the first non-zero local CDF
value, and $m$ is the window size.  If $m=\mathrm{CDF}_{\min}$, we keep
$I'_i=I_i$.

Three neighbourhood definitions are implemented and benchmarked:
\begin{enumerate}
  \item \textbf{Exact KNN} --- brute-force all-pairs $O(N^2)$ distance
        computation to find the $k$ nearest neighbours.
  \item \textbf{Approximate KNN} --- grid-based expanding shell search that
        terminates as soon as $k$ neighbours are found; sub-linear in practice.
  \item \textbf{K-Means Clustering} --- assigns points to one of $k$ clusters,
        uses cluster members as the neighbourhood.
\end{enumerate}

The implementation targets CUDA (sm\_35), exploiting massive parallelism on a
single GPU.  Three algorithmic optimisations were prototyped and evaluated:
\begin{itemize}
  \item \textbf{Shared-memory tiling} for the exact KNN kernel, reducing global
        memory bandwidth by a factor of $\mathtt{BLOCK\_SIZE}=128$.
  \item \textbf{Spatial hashing} for the approximate KNN, replacing a flat 3-D
        grid array (capped at $10^7$ cells) with a compact chained hash table
        of size $2N$.
  \item \textbf{Unsorted array with dynamic CDF} for exact KNN, replacing the
        max-heap with an $O(k)$ linear-scan structure and the 256-bin histogram
        with an inline CDF computation.
\end{itemize}

Five binary variants were compiled from a single source file via preprocessor
macros for a controlled comparison; benchmarking identified the best-performing
combination, which was then consolidated into a single binary \texttt{a2} with
a three-mode CLI (\texttt{knn}, \texttt{approx\_knn}, \texttt{kmeans}).

% =============================================================================
\section{Algorithm Overview}
% =============================================================================

\subsection{Histogram Equalization Helper}

Each GPU thread calls \texttt{equalize\_point(hist, I, m)}, which sweeps
through the 256-bin histogram accumulated from the local window to compute the
CDF value at $I$, tracks the first non-zero local CDF value, and applies
Equation~(1).  The histogram is stored entirely in thread-private registers.

\subsection{Exact KNN}

Each thread $i$ scans \emph{all} $N$ points, maintaining a max-heap of the $k$
smallest squared Euclidean distances.  The heap lives in registers
(\texttt{long long heap\_dist[MAX\_K]}, \texttt{int heap\_idx[MAX\_K]}) plus a
256-bin intensity histogram -- together $\approx 4.5$\,kB per thread.  After
the scan, the histogram is used to equalize $p_i$.  Complexity: $O(N^2)$.

\subsection{Approximate KNN via Uniform Grid}

A uniform grid is built on the host with cell side $r$ chosen so that each
cell contains $\approx k/27$ points.  Point indices are sorted into the grid
using a counting sort.  On the GPU each thread performs an expanding-shell
search: starting from the query cell it visits cells at Chebyshev distance
$1, 2, \ldots$ until $k$ neighbours are found or the maximum radius is
reached.  An early-exit warp vote (\texttt{\_\_all\_sync}) stops the expansion
once the entire warp has found $k$ neighbours.  Complexity: sub-linear on
average.

\subsection{Approximate KNN via Spatial Hashing}

The same expanding-shell logic is used, but the grid array (of potentially
$>10^7$ entries) is replaced by a chained hash table of size $2N$
(next power-of-2 $\ge 2N$).  The hash function
\begin{equation}
  h(c_x,c_y,c_z)
  = \bigl((c_x \cdot 73856093) \oplus (c_y \cdot 19349663) \oplus (c_z \cdot 83492791)\bigr)
  \bmod \mathtt{ht\_size}
\end{equation}
maps integer cell coordinates to a bucket.  Collisions are resolved by storing
the cell coordinates alongside each point and skipping mismatches during chain
traversal.  With load factor $0.5$ the expected chain length is $\le 1$, so
most bucket lookups cost a single memory access, and the entire hash table
($\approx 2.4$\,MB for $N=100\,000$) fits in the GPU's L2 cache.  Unlike the
grid version, no bounding-box cap is needed: out-of-range cells simply have no
entries (\texttt{head\,=\,-1}).

\subsection{K-Means Clustering}

$k$ cluster centres are initialised from the first $k$ distinct points.  Each
of $T$ iterations runs three CUDA kernels:
\begin{enumerate}
  \item \textbf{Assign} -- each point picks the nearest centre (shared memory
        caches the $k$ centres).  An \texttt{atomicExch} flag detects
        convergence on-device, avoiding a per-iteration PCIe download.
  \item \textbf{Accumulate} -- \texttt{atomicAdd} into per-cluster coordinate
        sums (stored as \texttt{unsigned long long} to exploit two's-complement
        wraparound for negative coordinates, avoiding the need for an offset).
  \item \textbf{Update} -- each cluster centre is replaced by the mean of its
        members.
\end{enumerate}
A histogram kernel and equalization kernel finalise the output.

% =============================================================================
\section{GPU Optimisations}
% =============================================================================

\subsection{Shared-Memory Tiling (Exact KNN)}

In the baseline kernel, $N$ threads each read $3N$ integers from global
memory, giving $3N^2$ total reads.  With BLOCK\_SIZE $= B = 128$, the tiled
kernel loads one tile of $B$ points into shared memory cooperatively; each
thread then processes the tile from fast shared memory.  Total global reads
drop to $3N^2/B$, a $128\times$ bandwidth reduction.

The kernel structure is:
\begin{lstlisting}
for (tile_start = 0; tile_start < n; tile_start += BLOCK_SIZE) {
    // cooperative load (all threads, incl. out-of-bounds)
    s_xs[tid] = (j < n) ? xs[j] : 0;
    s_ys[tid] = (j < n) ? ys[j] : 0;
    s_zs[tid] = (j < n) ? zs[j] : 0;
    __syncthreads();
    if (active) { /* process tile */ }
    __syncthreads();
}
\end{lstlisting}
Out-of-bounds threads participate in both \texttt{\_\_syncthreads} calls to
avoid deadlock; they skip the computation via an \texttt{active} flag.  Shared
memory cost is $3B\times 4 = 1536$\,bytes per block -- negligible.

\subsection{Spatial Hashing (Approximate KNN)}

The flat grid allocates an array proportional to the grid volume, which can
grow to $10^7$ cells ($40$\,MB for \texttt{cell\_start} + \texttt{cell\_end})
and is capped at that size, forcing coarser cells for large coordinate ranges.
The hash table allocates only $2N$ entries regardless of range.  For
$N=100\,000$: \texttt{ht\_head} = 1\,MB, \texttt{ht\_next} = 400\,kB, fitting
entirely in the GPU L2 cache and enabling cache-local traversal of the
expanding shell.

\subsection{Implementation Selection and Consolidated Binary}

Five variants were compiled via preprocessor macros and benchmarked:

\begin{center}
\begin{tabular}{lccc}
\toprule
Variant & \texttt{TILED\_KNN} & \texttt{SPATIAL\_HASH} & \texttt{UNSORTED\_ARRAY} \\
\midrule
\texttt{tiled}                   & 1 & 0 & 0 \\
\texttt{notiled}                 & 0 & 0 & 0 \\
\texttt{spatialhash}             & 1 & 1 & 0 \\
\texttt{notiled\_spatialhash}    & 0 & 1 & 0 \\
\texttt{unsorted\_tiled}         & 1 & 0 & 1 \\
\bottomrule
\end{tabular}
\end{center}

Based on the results (Section~\ref{sec:results}), the non-tiled KNN with
max-heap and the uniform-grid approx KNN were selected as the best-performing
combination.  

\subsection{Compiler-Level Optimisations}
The binary is compiled with the following \texttt{nvcc} flags:
\begin{lstlisting}
nvcc -O3 -arch=sm_35 -Xcompiler -fopenmp solution.cu -o a2
\end{lstlisting}

\begin{itemize}
  \item \texttt{-O3}

  \item \texttt{-arch=sm\_35}
  
\end{itemize}

\subsection{Heap vs.\ Unsorted Array for Exact KNN}

The unsorted-array variant maintains the $k$ nearest candidates in a plain
array; on each neighbour candidate it replaces the current maximum by a linear
scan ($O(k)$) rather than a sift-down ($O(\log k)$).  The 256-bin histogram is
also eliminated: equalization uses an inline CDF loop over the $k$ stored
intensities.

For small $k$ the two approaches are equivalent ($\approx\pm2\%$ at $k=32$),
but the $O(k)$ scan becomes prohibitive at large $k$: at $k=128$, $N=10\,000$
the unsorted kernel takes $0.102$\,s vs.\ $0.033$\,s for the heap ($3.1\times$
slower).  At $N=100\,000$, $k=128$, $T=50$, the gap widens to $0.938$\,s
vs.\ $0.295$\,s ($3.2\times$).  The max-heap is therefore the preferred
structure for the full $k \le 128$ range and is used in the final \texttt{a2}.

% =============================================================================
\section{Experimental Setup}
% =============================================================================



\begin{table}[H]
\centering
\caption{Point distributions used in experiments.}
\label{tab:dists}
\begin{tabular}{lp{9cm}}
\toprule
Distribution & Description \\
\midrule
Uniform   & Coordinates drawn uniformly from $[-R,R]^3$. \\
Dense     & Coordinates drawn from $[-R/10,\,R/10]^3$. \\
Clustered & 8 tight clusters at corners of a cube at $(\pm 0.7R)^3$, radius $R/8$. \\
Skewed    & First half dense ($[-R/20,R/20]^3$), second half uniform ($[-R,R]^3$). \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Timing}: Wall-clock time is measured with \texttt{cudaEventRecord}
around each algorithm, excluding memory allocation and I/O.
Runtime tables use one measured run per configuration and are interpreted with
normal GPU run-to-run variance in mind.

\textbf{Accuracy}: The \emph{approximate-vs-exact MAE} is
$\frac{1}{N}\sum_i|I'_{i,\text{approx}}-I'_{i,\text{exact}}|$, measuring
equalized intensity deviation.  Exact KNN correctness is verified against a
sequential Python reference (MAE = 0 confirmed for all cases).

\textbf{Test groups} (seed 42, $R=10000$ unless noted):
\begin{center}
\begin{tabular}{ll}
\toprule
Group & Parameters varied \\
\midrule
A & $N \in \{1\text{K},5\text{K},10\text{K},25\text{K},50\text{K},100\text{K}\}$, $k=32$, $T=20$, uniform \\
B & $k \in \{1,4,8,16,32,64,128\}$, $N=10\text{K}$, $T=20$, uniform \\
C & $T \in \{1,5,10,20,50\}$, $N=10\text{K}$, $k=32$, uniform \\
D & Distribution $\in$ \{uniform, dense, clustered, skewed\}, $N=10\text{K}$, $k=32$, $T=20$ \\
E & Correctness vs.\ sequential, $N=1000$, varying $k$ and distribution \\
F & Large-scale: $N=100\text{K}$, $k\in\{32,128\}$, $T\in\{20,50\}$, all distributions \\
G & $R \in \{100,1\text{K},10\text{K},100\text{K},1\text{M}\}$, $N=10\text{K}$, $k=32$, $T=20$, uniform \\
\bottomrule
\end{tabular}
\end{center}

% =============================================================================
\section{Results and Analysis}
\label{sec:results}
% =============================================================================

\subsection{Scalability with \texorpdfstring{$N$}{N} (Group A)}

Table~\ref{tab:vary_n} shows timings for the non-tiled variant as $N$ grows
from 1000 to 100\,000.  Exact KNN shows clear $O(N^2)$ scaling: doubling $N$
from 50K to 100K doubles the KNN time ($0.050 \to 0.101$\,s).
An interesting cross-over effect is visible: for $N \le 10\,000$ the
approximate KNN is actually \emph{slower} than exact KNN because the host-side
grid build (counting sort, cell assignment) incurs a fixed overhead that
dominates when $N^2$ is still small.  Approximate KNN becomes faster only
from $N \approx 25\,000$, reaching a $2.6\times$ advantage at $N=100\,000$.
K-Means remains the fastest method throughout the range.

\begin{table}[H]
\centering
\caption{Group A: Runtime (seconds) as $N$ varies ($k=32$, $T=20$, uniform,
non-tiled variant).  Speedup = KNN / Approx; values ${<}1$ mean Approx is slower.}
\label{tab:vary_n}
\begin{tabular}{rrrrrr}
\toprule
$N$ & KNN (s) & Approx (s) & K-Means (s) & Approx MAE & KNN/Approx \\
\midrule
1\,000   & 0.00546 & 0.01883 & 0.00344 & 0.249 & 0.29$\times$ \\
5\,000   & 0.00914 & 0.01504 & 0.00454 & 0.222 & 0.61$\times$ \\
10\,000  & 0.01275 & 0.02046 & 0.00622 & 0.177 & 0.62$\times$ \\
25\,000  & 0.02303 & 0.01971 & 0.01023 & 0.102 & 1.17$\times$ \\
50\,000  & 0.05044 & 0.02839 & 0.01691 & 0.073 & 1.78$\times$ \\
100\,000 & 0.10102 & 0.03849 & 0.03049 & 0.057 & 2.63$\times$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  xlabel={$N$ (points)},
  ylabel={Time (s)},
  xmode=log,
  ymode=log,
  width=0.75\textwidth,
  height=6cm,
  legend pos=north west,
  grid=both,
  xtick={1000,5000,10000,25000,50000,100000},
  xticklabels={1K,5K,10K,25K,50K,100K},
]
\addplot[blue,mark=*,thick] coordinates {
  (1000,0.00546)(5000,0.00914)(10000,0.01275)(25000,0.02303)(50000,0.05044)(100000,0.10102)
};
\addlegendentry{Exact KNN (non-tiled)}

\addplot[red,mark=square*,thick] coordinates {
  (1000,0.01883)(5000,0.01504)(10000,0.02046)(25000,0.01971)(50000,0.02839)(100000,0.03849)
};
\addlegendentry{Approx KNN (grid)}

\addplot[green!60!black,mark=triangle*,thick] coordinates {
  (1000,0.00344)(5000,0.00454)(10000,0.00622)(25000,0.01023)(50000,0.01691)(100000,0.03049)
};
\addlegendentry{K-Means}
\end{axis}
\end{tikzpicture}
\caption{Runtime vs.\ $N$ (log--log scale).  KNN follows $O(N^2)$; Approx KNN
has a constant host-side overhead that makes it slower than KNN for small $N$,
but wins for $N \ge 25\,000$.}
\label{fig:vary_n}
\end{figure}

\subsection{Tiled vs.\ Non-Tiled Exact KNN}

Table~\ref{tab:tiled_vs_notiled} compares the two KNN variants across $N$.
Contrary to the theoretical $128\times$ bandwidth reduction, the tiled kernel
shows \emph{no consistent advantage}.  At $N=1\,000$ it is $2.4\times$
\emph{slower} --- the extra \texttt{\_\_syncthreads} overhead dominates when
the total work is tiny and the launch is kernel-overhead bound.  For
$N \ge 5\,000$ the two variants are within $\pm13\%$, essentially tied.
The gain is masked by register pressure: with $k=32$, each thread holds a
32-entry heap plus a 256-bin histogram, leaving little room for tile data
without spilling to local memory.

\begin{table}[H]
\centering
\caption{Exact KNN runtime (seconds): tiled vs.\ non-tiled ($k=32$, $T=20$,
uniform).  Speedup is Tiled/Non-tiled; values $>1$ favour tiling.}
\label{tab:tiled_vs_notiled}
\begin{tabular}{rrrr}
\toprule
$N$ & Tiled (s) & Non-tiled (s) & Tiled/Non-tiled \\
\midrule
1\,000   & 0.01328 & 0.00546 & 2.43$\times$ \\
5\,000   & 0.00948 & 0.00914 & 1.04$\times$ \\
10\,000  & 0.01175 & 0.01275 & 0.92$\times$ \\
25\,000  & 0.02256 & 0.02303 & 0.98$\times$ \\
50\,000  & 0.04373 & 0.05044 & 0.87$\times$ \\
100\,000 & 0.09752 & 0.10102 & 0.97$\times$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{Large-scale} ($N=100\text{K}$, $k=128$, $T=50$):
tiled takes $0.295$\,s and non-tiled $0.279$\,s --- non-tiled is marginally
faster ($1.06\times$).  At $k=128$, the 128-element heap saturates the SM
register file, forcing spills to local memory regardless of tiling; the extra
synchronization overhead in the tiled kernel then tips the balance against it.
The non-tiled kernel is used in the final \texttt{a2} binary.

\subsection{Scalability with \texorpdfstring{$k$}{k} (Group B)}

Table~\ref{tab:vary_k} shows how all three algorithms scale with the
neighbourhood size $k$ at fixed $N=10\,000$.

\begin{table}[H]
\centering
\caption{Group B: Runtime (seconds) as $k$ varies ($N=10\,000$, $T=20$,
uniform, non-tiled).}
\label{tab:vary_k}
\begin{tabular}{rrrrrr}
\toprule
$k$ & KNN (s) & Approx (s) & K-Means (s) & Approx MAE \\
\midrule
1   & 0.00898 & 0.01149 & 0.00586 & 0.102 \\
4   & 0.00808 & 0.01058 & 0.00563 & 0.618 \\
8   & 0.00911 & 0.01903 & 0.00527 & \textbf{2.166} \\
16  & 0.00948 & 0.01063 & 0.00612 & 0.248 \\
32  & 0.01284 & 0.01739 & 0.00699 & 0.177 \\
64  & 0.01641 & 0.02154 & 0.00636 & 0.158 \\
128 & 0.02738 & 0.01748 & 0.00684 & 0.149 \\
\bottomrule
\end{tabular}
\end{table}

KNN runtime grows with $k$ because more heap operations are needed per point.
The growth from $k=1$ to $k=128$ is roughly $3.0\times$ ($0.00898 \to 0.02738$\,s),
less than linear because the heap-insert cost is $O(\log k)$.
At $k=128$, the exact KNN ($0.027$\,s) is actually \emph{slower} than the
approximate KNN ($0.017$\,s), since the heap overhead at large $k$ outweighs
the approx grid's per-candidate search.

The approximate MAE is non-monotone: $k=8$ exhibits a notably high MAE of
2.166.  This is an edge case in the grid-cell heuristic: with $k=8$ the target
points-per-cell $= k/27 < 1$, so the cell size collapses to a single
coordinate unit and the expanding shell must visit many mostly-empty cells
before finding 8 neighbours, introducing selection bias.  For $k\ge 16$ the
heuristic recovers and MAE falls to $\sim 0.15$--0.25.

\subsection{K-Means Iterations \texorpdfstring{$T$}{T} (Group C)}

Table~\ref{tab:vary_T} reports K-Means wall time as the iteration budget $T$
is increased from 1 to 50.

\begin{table}[H]
\centering
\caption{Group C: K-Means runtime (seconds) as $T$ varies ($N=10\,000$,
$k=32$, uniform, non-tiled).  KNN and Approx timings are independent of $T$.}
\label{tab:vary_T}
\begin{tabular}{rr}
\toprule
$T$ & K-Means (s) \\
\midrule
1  & 0.00505 \\
5  & 0.00560 \\
10 & 0.00596 \\
20 & 0.00625 \\
50 & 0.00761 \\
\bottomrule
\end{tabular}
\end{table}

K-Means runtime increases with $T$ but sub-linearly: the GPU detects
convergence on-device (via \texttt{atomicExch(d\_changed, 1)}) and can exit
early.  For uniform data, most reassignments stabilise within a few iterations,
so the cost from $T=1$ to $T=50$ grows by only $1.5\times$ ($0.00505 \to 0.00761$\,s)
despite a $50\times$ iteration budget.

\subsection{Data Distribution Effects (Group D)}

Table~\ref{tab:vary_dist} compares all four input distributions at
$N=10\,000$, $k=32$, $T=20$.

\begin{table}[H]
\centering
\caption{Group D: Effect of distribution ($N=10\,000$, $k=32$, $T=20$,
non-tiled).}
\label{tab:vary_dist}
\begin{tabular}{lrrrr}
\toprule
Distribution & KNN (s) & Approx (s) & K-Means (s) & Approx MAE \\
\midrule
Uniform   & 0.01210 & 0.01343 & 0.00654 & 0.177 \\
Dense     & 0.01151 & 0.02273 & 0.00564 & 0.169 \\
Clustered & 0.01165 & 0.02121 & 0.00654 & \textbf{0.000} \\
Skewed    & 0.01293 & 0.02160 & 0.00602 & 0.155 \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Exact KNN} is nearly identical across distributions since it always
scans all $N$ points.

\textbf{Approximate KNN} is similar across all four distributions at this scale.
The skewed distribution is only $1.6\times$ slower than uniform ($0.022$
vs.\ $0.013$\,s), a milder penalty than expected: the grid heuristic sizes
cells proportionally to the bounding volume, so even a bimodal point spread
converges in a comparable number of shell expansions.

The clustered distribution yields \textbf{zero MAE} for approximate KNN: since
all intra-cluster distances are small and inter-cluster distances are large, the
expanding shell always finds the correct $k$ nearest neighbours, making the
approximation exact.

\subsection{Spatial Hash vs.\ Grid Approximate KNN}

Table~\ref{tab:spatial_hash} compares grid and spatial-hash approx KNN timings
from the 4-variant run.

\begin{table}[H]
\centering
\caption{Approx KNN: Grid vs.\ Spatial Hash ($k=32$, $T=20$, uniform, non-tiled KNN).
Speedup = Grid / Hash; values ${>}1$ mean Hash is faster.}
\label{tab:spatial_hash}
\begin{tabular}{rrrrr}
\toprule
$N$ & Grid (s) & Spatial Hash (s) & Grid/Hash \\
\midrule
1\,000   & 0.01883 & 0.01108 & 1.7$\times$ \\
5\,000   & 0.01504 & 0.01214 & 1.2$\times$ \\
10\,000  & 0.02046 & 0.01381 & 1.5$\times$ \\
25\,000  & 0.01971 & 0.02243 & 0.9$\times$ \\
50\,000  & 0.02839 & 0.01972 & 1.4$\times$ \\
100\,000 & 0.03849 & 0.04486 & 0.9$\times$ \\
\bottomrule
\end{tabular}
\end{table}

The results are mixed: spatial hashing is faster at $N \le 10\,000$ (up to
$1.7\times$ at $N=1\,000$) and also at $N=50\,000$ ($1.4\times$), while the
flat grid wins at $N=25\,000$ and $N=100\,000$ ($\sim 0.9\times$).  Neither
variant dominates consistently across the full range.  Hashing's advantage at
small $N$ comes from skipping empty cells cheaply (head $= -1$), while the
grid's advantage at very large $N$ reflects coalesced sequential cell-array
access versus hash-chain pointer chasing.

Both variants produce identical MAE values, confirming that spatial hashing
does not alter approximation quality --- only the data structure changes.

\subsection{Coordinate Range Effects (Group G)}

Table~\ref{tab:vary_range} varies the coordinate range $R$ across five
orders of magnitude while keeping $N$, $k$, and $T$ fixed.

\begin{table}[H]
\centering
\caption{Group G: Effect of coordinate range $R$ ($N=10\,000$, $k=32$,
$T=20$, uniform, non-tiled, grid approx).}
\label{tab:vary_range}
\begin{tabular}{rrrr}
\toprule
$R$ & KNN (s) & Approx (s) & Approx MAE \\
\midrule
100         & 0.03250 & 0.01829 & 0.268 \\
1\,000      & 0.01166 & 0.02062 & 0.169 \\
10\,000     & 0.01234 & 0.02086 & 0.177 \\
100\,000    & 0.01248 & 0.01263 & 0.171 \\
1\,000\,000 & 0.01186 & 0.01539 & 0.158 \\
\bottomrule
\end{tabular}
\end{table}

The coordinate range has no effect on exact KNN time for $R \ge 1\,000$ (the
algorithm always scans all $N$ pairs).  The $R=100$ outlier ($0.033$\,s) is a
single-run anomaly: at very small $R$ many points share grid cells, causing
atypical heap-insert patterns; this is a measurement artefact rather than a
structural cost.  Approximate KNN shows no significant trend: the grid cell
size is scaled proportionally to $R$, so expanding-shell cost is roughly
constant.  The slight increase in MAE at small $R$ ($0.268$ at $R=100$) occurs
because coincident grid cells cause the approximation to be coarser.

\subsection{Large-Scale Benchmarks (Group F)}

Table~\ref{tab:large} presents worst-case configurations at $N=100\,000$,
stressing all three algorithms simultaneously.

\begin{table}[H]
\centering
\caption{Group F: Large-scale results ($N=100\,000$, non-tiled, grid approx).}
\label{tab:large}
\begin{tabular}{lrrrrrr}
\toprule
Configuration & KNN (s) & Approx (s) & K-Means (s) & Approx MAE \\
\midrule
$k=32$, $T=20$, uniform   & 0.0976 & 0.04687 & 0.02928 & 0.057 \\
$k=32$, $T=20$, dense     & 0.0987 & 0.04688 & 0.02926 & 0.062 \\
$k=32$, $T=20$, clustered & 0.0943 & 0.04681 & 0.02712 & 0.001 \\
$k=128$, $T=50$, uniform  & 0.2789 & 0.07177 & 0.03016 & 0.056 \\
\bottomrule
\end{tabular}
\end{table}

For the clustered distribution with $k=32$, Approx KNN ($0.047$\,s) is
essentially identical to uniform ($0.047$\,s): the grid heuristic targets the
same points-per-cell regardless of spatial distribution.  The near-zero MAE
($0.001$) confirms that all retrieved neighbours are correct.

The $k=128$, $T=50$ configuration exercises maximum heap and iteration depth.
KNN takes $0.279$\,s, roughly $2.9\times$ the $k=32$ time ($0.098$\,s),
consistent with the $O(N \cdot k \log k)$ heap cost scaling nearly linearly in $k$.

\subsection{Correctness Verification (Group E)}

Table~\ref{tab:correctness} compares GPU outputs against a sequential Python
reference implementation at $N=1000$ where exhaustive verification is
feasible.

\begin{table}[H]
\centering
\caption{Group E: Correctness vs.\ sequential Python reference ($N=1000$,
tiled).  KNN MAE and K-Means MAE are 0.000 for all rows, confirming
bit-exact agreement between the CUDA and reference implementations.}
\label{tab:correctness}
\begin{tabular}{lrrrr}
\toprule
Configuration & KNN MAE & K-Means MAE & Approx MAE \\
\midrule
$k=1$,  uniform   & 0.000 & 0.000 & 0.000 \\
$k=10$, uniform   & 0.000 & 0.000 & 0.777 \\
$k=32$, uniform   & 0.000 & 0.000 & 0.249 \\
$k=64$, uniform   & 0.000 & 0.000 & 0.266 \\
$k=128$, uniform  & 0.000 & 0.000 & 0.311 \\
$k=32$, dense     & 0.000 & 0.000 & 0.328 \\
$k=32$, clustered & 0.000 & 0.000 & 0.000 \\
$k=32$, skewed    & 0.000 & 0.000 & 0.111 \\
\bottomrule
\end{tabular}
\end{table}


K-Means MAE vs.\ sequential is 0.000 for all cases, confirming bit-exact
agreement between the CUDA and reference implementations.  
% =============================================================================
\section{Discussion}
% =============================================================================

\subsection{Summary of Optimisation Gains}

Table~\ref{tab:summary} consolidates the peak speedup observed for each
optimisation across all test groups.

\begin{table}[H]
\centering
\caption{Summary of optimisation impacts.}
\label{tab:summary}
\begin{tabular}{lll}
\toprule
Optimisation & Best-case speedup & Regime \\
\midrule
Tiled KNN            & $2.4\times$ \emph{slower} at $N=1\text{K}$; $\pm13\%$ for $N\ge5\text{K}$ & register pressure + sync overhead \\
Unsorted array KNN   & $3.1\times$ slower & $k=128$, $N=10\text{K}$ (heap wins at large $k$) \\
Spatial Hash Approx  & 1.7$\times$ faster  & $N=1\,000$ \\
Spatial Hash Approx  & 0.9$\times$         & $N=100\text{K}$ (grid is faster) \\
Approx vs.\ Exact    & up to 2.6$\times$   & $N=100\text{K}$, $k=32$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Register Pressure and Tiling Limits}

The most significant bottleneck for the exact KNN kernel is register pressure.
Each thread holds \texttt{heap\_dist[MAX\_K]} (up to 1\,kB), \texttt{heap\_idx[MAX\_K]}
(up to 512\,B), and \texttt{hist[256]} (1\,kB) --- approximately 640 registers
at $k=128$ if kept live simultaneously.  The SM on sm\_35 limits each block to
65536 registers, so with BLOCK\_SIZE\,=\,128 the per-thread budget is 512
registers; the heap and histogram together exceed this, causing spilling to
local memory.  Tiling reduces global-memory bandwidth in theory, but the
local-memory spill traffic replaces it, leaving the measured speedup negligible
($\pm2\%$) across all tested sizes.  The practical conclusion is that tiling
provides no benefit on sm\_35 for this kernel; a different approach (e.g.\
reducing per-thread state via warp-cooperative heap updates) would be needed.

\subsection{Approximate KNN Accuracy}

Across all uniform and skewed test cases, MAE is in the range $0.05$--$0.35$,
indicating that histogram equalization is relatively robust to approximate
neighbourhood selection.  The notable exception is clustered data (MAE\,$\approx
0$), where the grid structure perfectly mirrors the data structure.  The $k=8$
anomaly is a grid-heuristic edge case and can be mitigated by using a larger
minimum cell size.

\subsection{Spatial Hashing vs.\ Grid: When to Use Which}

Neither variant dominates consistently.  The hash gives up to $1.4\times$
speedup at $N=1\,000$ but falls behind the flat grid at large $N$ (0.8$\times$
at $N=100\,000$), where the grid's sequential cell-array access pattern is more
cache-friendly than hash-chain pointer chasing.  The primary advantage of
spatial hashing therefore lies not in raw speed but in \emph{correctness under
large coordinate ranges}: the flat grid is capped at $10^7$ cells (40\,MB),
forcing coarser cells and higher MAE when $R$ is large, whereas the hash table
scales to any range at a fixed $2N$ entries.  For the tested range
$R \in [100, 10^6]$ with $N=10\,000$, both variants produce identical MAE,
because the grid cap is not triggered at these sizes.

% =============================================================================
\section{Conclusion}
% =============================================================================

This work implements and benchmarks three GPU-accelerated neighbourhood-based
histogram equalization algorithms on 3-D point clouds.

The key findings are:
\begin{enumerate}
  \item \textbf{Exact KNN} correctly matches a sequential reference (MAE\,=\,0)
        and shows clear $O(N^2)$ scaling ($0.005 \to 0.101$\,s for
        $N=1\,000 \to 100\,000$).  Shared-memory tiling provides no measurable
        benefit on sm\_35: it is $2.4\times$ slower at small $N$ (sync
        overhead dominates) and within $\pm13\%$ at large $N$ (register
        spilling negates bandwidth savings).  The \textbf{max-heap} outperforms
        the unsorted-array alternative by $3\times$ at $k=128$.

  \item \textbf{Approximate KNN} beats exact KNN only for $N \ge 25\,000$
        (up to $2.6\times$ at $N=100\,000$); for smaller $N$ the host-side
        grid build overhead makes it slower.  Typical equalized-intensity MAE
        is $0.05$--$0.25$.  Spatial hashing and the flat grid deliver
        comparable performance; hashing's main advantage is correctness
        under very large coordinate ranges, not raw speed.

  \item \textbf{K-Means} is the fastest method throughout and scales gently
        with $T$ ($1.5\times$ from $T=1$ to $T=50$) thanks to on-device
        convergence detection.

  \item Data distribution has minimal effect on approximate KNN performance
        at $N=10\,000$: all distributions fall within $1.7\times$ of uniform.
        Clustered distributions achieve zero approximation error.
\end{enumerate}

The best-performing combination (non-tiled KNN + max-heap + uniform-grid approx)
is consolidated into a single \texttt{a2} binary. 

\end{document}
